---
title: "README"
output: html_document
date: "2024-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# FreSpeD

**Frequency specific change point detection (FreSpeD)** is a method proposed by [Schröder & Ombao, 2019](https://doi.org/10.1080/01621459.2018.1476238) for detecting change points in EEG signals. By locally estimating spectral density matrices and applying CUSUM statistics, FreSpeD efficiently identifies significant changes in the spectral profile. The method has demonstrated speed and effectiveness in detecting epileptic seizures.

This package was initially published on CRAN in September 2015 and is compatible with R version 3.2.2. However, it has since been archived and is no longer available for versions of R released after 3.2.2. The purpose of this repository is to provide a reimplementation of the original package and introduce new functionalities.


## Installation

You can install the development version of <tt>FreSpeD</tt> from
[GitHub](https://github.com) with:

```{r}
devtools::install_github("HoussamBoukhe/FreSpeD")
```

## Dependencies
R version: 4.4.1

doParallel 1.0.17

foreach 1.5.2

reshape 1.4.4

ggplot2 3.5.1

# Examples

We give some examples of application of the FreSpeD method to simulated data,
then to datasets from [Physionet](https://physionet.org/).

For simulated data, we use a mixture of AR(2) models. This can be done as 
follows 


```{r}
library(FreSpeD)
library(doParallel)
library(foreach)
library(ggplot2)
library(reshape2)
library(eegkit)
```

```{r echo=FALSE}
set.seed(2024)

```

```{r}
# 1 --- Generating synthetic AR(2) models
f_sampling <- 256 # Sampling rate 

magnitude <- 1.2
phase_delta <- 2/f_sampling
phase_theta <- 6/f_sampling
phase_alpha <- 10/f_sampling
phase_beta <- 21/f_sampling
phase_gamma <- 40/f_sampling


model_delta <- list(ar = c(2/magnitude * cos(2*pi*phase_delta), -1/magnitude^2))
model_theta <- list(ar = c(2/magnitude * cos(2*pi*phase_theta), -1/magnitude^2))
model_alpha <- list(ar = c(2/magnitude * cos(2*pi*phase_alpha), -1/magnitude^2))
model_beta <- list(ar = c(2/magnitude * cos(2*pi*phase_beta), -1/magnitude^2))
model_gamma <- list(ar = c(2/magnitude * cos(2*pi*phase_gamma), -1/magnitude^2))

```


The above parametrization of the AR(2) model is used to be able to choose the 
peack of the generated model around a given frequency. For more details see
[Hernando Ombao and Marco Pinto. Spectral dependence](https://www.sciencedirect.com/science/article/pii/S2452306222001101)

```{r}
# Generating the building blocks for the synthetic data 
Z_delta <- arima.sim(model = model_delta, n = 256*60)
Z_theta <- arima.sim(model = model_theta, n = 256*60)
Z_alpha <- arima.sim(model = model_alpha, n = 256*60)
Z_beta <- arima.sim(model = model_beta, n = 256*60)
Z_gamma <- arima.sim(model = model_gamma, n= 256*60)

spec_density_delta <- spec.pgram(Z_delta, log = 'no', plot = FALSE)
spec_density_theta <- spec.pgram(Z_theta, log = 'no', plot = FALSE)
spec_density_alpha <- spec.pgram(Z_alpha, log = 'no', plot = FALSE)
spec_density_beta <- spec.pgram(Z_beta, log = 'no', plot = FALSE)
spec_density_gamma <- spec.pgram(Z_gamma, log = 'no', plot = FALSE)

# Multiply frequencies by the sampling rate
frequencies_hz <- spec_density_alpha$freq * f_sampling

# Plot the periodogram with adjusted frequencies
plot(frequencies_hz, spec_density_delta$spec, type = "l", xlab = "Frequency (Hz)", ylab = "Spectral Density", col = "red")
plot(frequencies_hz, spec_density_theta$spec, col = "blue", type ='l', xlab = "Frequency (Hz)", ylab = "Spectral Density")
plot(frequencies_hz, spec_density_alpha$spec, col = "green", type ='l', xlab = "Frequency (Hz)", ylab = "Spectral Density")
plot(frequencies_hz, spec_density_beta$spec, col = "orange", type = 'l', xlab = "Frequency (Hz)", ylab = "Spectral Density")
plot(frequencies_hz, spec_density_gamma$spec, col = "purple", type = 'l', xlab = "Frequency (Hz)", ylab = "Spectral Density")

```
The support of each model has to be in:

delta: 0.1-5 Hz
theta: 5-8 Hz
alpha: 8-12 Hz
beta: 12-30 Hz
gamma: 30-50 Hz,

so we apply a linear filter to cut the supports of the spectral densities

```{r}

Z_delta <- eegfilter(Z_delta, Fs = f_sampling, lower = 1, upper = 4, method = "butter", order = 4)
Z_theta <- eegfilter(Z_theta, Fs = f_sampling, lower = 4, upper = 8, method = "butter", order = 4)
Z_alpha <- eegfilter(Z_alpha, Fs = f_sampling, lower = 8, upper = 12, method = "butter", order = 4)
Z_beta <- eegfilter(Z_beta, Fs = f_sampling, lower = 13, upper = 30, method = "butter", order = 4)
Z_gamma <- eegfilter(Z_gamma, Fs = f_sampling, lower = 30, upper = 50, method = "butter", order = 4)

spec_density_delta <- spec.pgram(Z_delta, log = 'no', plot = FALSE)
spec_density_theta <- spec.pgram(Z_theta, log = 'no', plot = FALSE)
spec_density_alpha <- spec.pgram(Z_alpha, log = 'no', plot = FALSE)
spec_density_beta <- spec.pgram(Z_beta, log = 'no', plot = FALSE)
spec_density_gamma <- spec.pgram(Z_gamma, log = 'no', plot = FALSE)

# Multiply frequencies by the sampling rate
frequencies_hz <- spec_density_alpha$freq * f_sampling

# Plot the periodogram with adjusted frequencies
plot(frequencies_hz, spec_density_delta$spec, type = "l", xlab = "Frequency (Hz)", ylab = "Spectral Density", col = "red")
plot(frequencies_hz, spec_density_theta$spec, col = "blue", type ='l', xlab = "Frequency (Hz)", ylab = "Spectral Density")
plot(frequencies_hz, spec_density_alpha$spec, col = "green", type ='l', xlab = "Frequency (Hz)", ylab = "Spectral Density")
plot(frequencies_hz, spec_density_beta$spec, col = "orange", type = 'l', xlab = "Frequency (Hz)", ylab = "Spectral Density")
plot(frequencies_hz, spec_density_gamma$spec, col = "purple", type = 'l', xlab = "Frequency (Hz)", ylab = "Spectral Density")

```
Now, we use the building blocks to generate some examples, then apply the FreSpeD.
In the following, we generate a bivariate time series X, with a change point in 
the middle of this time series.

```{r}
Z <- t(cbind(Z_delta,Z_theta,Z_alpha,Z_beta,Z_gamma))

Mat1 <- matrix(data = c(1,1,0,0,0, 0,0,0,1,1), nrow = 2, ncol = 5, byrow = TRUE)
Mat2 <- matrix(data = c(1,1,1,0,0, 0,0,1,1,1), nrow = 2, ncol = 5, byrow = TRUE)

# Generate a two dimensional time series, its first component contains 
# 'delta and theta' frequencies, and its second component contains 'beta and gamma'
X1 <- t(Mat1%*%Z)

# Generate a two dimensional time series, its first component contains 
# 'delta, theta and alpha' frequencies, and its second component contains 
# 'alpha, beta and gamma' 
X2 <- t(Mat2%*%Z)

X <- rbind(X1,X2)

```


```{r}
# Time variying spectrum of the components of X2
FreSpeD:::tv_spectrum(X2[,1], windowLen = 512, plot = TRUE)
FreSpeD:::tv_spectrum(X2[,2], windowLen = 512, plot = TRUE)

```

```{r}
# Cross-coherence between the channels of X1 and X2
FreSpeD:::tv_coherence(X1, windowLen = 512, M_welch = 2,plot = TRUE, transform = "",
            bands_analysis = TRUE, f_sampling = f_sampling)
FreSpeD:::tv_coherence(X2, windowLen = 512, M_welch = 2, plot = TRUE, transform = "",
            bands_analysis = TRUE, f_sampling = f_sampling)

```
Notice the the cross-coherenc between the channels of X2 is 'darker' since its 
two components share a common alpha frequency.

```{r}

cp <- FreSpeD:::tv_estimations_cp(X, windowLen = 512, M_welch = 2,plot = FALSE, bands_analysis = TRUE, f_sampling = f_sampling) 
summary <- FreSpeD:::FreSpeD_summary(cp, plot = FALSE, windowLen = 512)

```

The length of the time series X is 30720, and the change occurs at 15360.

```{r}
summary$`Change points over time`
```
## Application to real EEG datasets
In this section, we use datasets from Physionet to test the FreSpeD algorithm and conduct analyses using it.
```{r}
# This chunk defines a functions that read an EDF file, then returns the time 
# series as a matrix
install.packages('edfReader')
library(edfReader)
# Specify the path to your EDF file
edf_get_signals <-function(edf_file_path, nbr_channels = NULL){ 
  
  edf_header <- readEdfHeader(edf_file_path)

  if (is.null(nbr_channels)) {
    nbr_channels <- edf_header$nSignals
  }
  
  edf_signals <- readEdfSignals(edf_header)
  signals_list <- list()
  
  # Loop over all channels and store the signals in the list
  for (i in seq_along(edf_signals)) {
    signals_list[[i]] <- edf_signals[[i]]$signal
  }
  
  # Combine the list of signals into a matrix
  signal_matrix <- do.call(cbind, signals_list)
  
  signal_matrix <- signal_matrix[,1:nbr_channels]
  return(signal_matrix)
}

```
### Application to EEG During Mental Arithmetic Tasks
In this study, we analyzed 10 patients (the original dataset contain 36 patient), 
each with two recordings: one during rest and another while performing a mental arithmetic task. We employed the FreSpeD method to calculate the total number of change points in each recording and subsequently compared the total number of change points observed during rest versus task performance.

```{r}


```



